---
title: "EGU 2020 blog post: 'Accessing environmental time series data in R from Sensor Observation Services with ease'"
author:
  name: "Daniel Nüst¹, Eike H. Jürrens², Benedikt Gräler², and Simon Jirka²"
  affiliation: "¹: University of Münster, Institute for Geoinformatics, Opening Reproducible Research (o2r), Münster, Germany (daniel.nuest@uni-muenster.de); ²: Institute for Geoinformatics, University of Münster, Germany. 52°North GmbH, Münster, Germany."
  email: "daniel.nuest@uni-muenster.de, e.h.juerrens@52north.org, b.graeler@52north.org, and s.jirka@52north.org"
output:
  rmarkdown::html_vignette
abstract: >
  Time series data of in-situ measurements is the key to many environmental studies.
  The first challenge in any analysis typically arises when the data needs to be imported into the analysis framework.
  Standardisation is one way to lower this burden.
  Unfortunately, relevant interoperability standards might be challenging for non-IT experts as long as they are not dealt with behind the scenes of a client application.
  One standard to provide access to environmental time series data is the Sensor Observation Service ([SOS](https://en.wikipedia.org/wiki/Sensor_Observation_Service)) specification published by the Open Geospatial Consortium ([OGC](https://www.ogc.org/)). SOS instances are currently used in a broad range of applications such as hydrology, air quality monitoring, and ocean sciences. Data sets provided via an SOS interface can be found around the globe from Europe to New Zealand.
  
  The R package sos4R (Nüst et al., 2011) is an extension package for the R environment for statistical computing and visualization (https://www.r-project.org/), which has been demonstrated a a powerful tools for conducting and communicating geospatial research (cf. Pebesma et al., 2012; ).
  sos4R comprises a client that can connect to an SOS server.
  The user can use it to query data from SOS instances using simple R function calls.
  It provides a convenience layer for R users to integrate observation data from data access servers compliant with the SOS standard without any knowledge about the underlying technical standards.
  To further improve the usability for non-SOS experts, a recent update to sos4R includes a set of wrapper functions, which remove complexity and technical language specific to OGC specifications.
  This update also features specific consideration of the OGC SOS 2.0 Hydrology Profile and thereby opens up a new scientific domain.

  In our presentation we illustrate use cases and examples building upon sos4R easing the access of time series data in an R and Shiny (https://shiny.rstudio.com/) context.
  We demonstrate how the abstraction provided in the client library makes sensor observation data for accessible and further show how sos4R allows the seamless integration of distributed observations data, i.e., across organisational boundaries, into transparent and reproducible data analysis workflows.
vignette: >
  %\VignetteIndexEntry{5. Convenient wrapper functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r blog_post, include = FALSE, eval=FALSE}
# run this code chunk to render an HTML snippet for use in Wordpress
library("rmarkdown")
render("sos4R-vignette-10-egu-2020.Rmd",
       output_format = html_fragment(
         #includes = includes(before_body = "javascripts.html") # TODO add JS loading here?
       ),
       # does not work yet because it is missing the JavaScript
       #output_format = html_document(self_contained = TRUE),
       output_file = "copy-to-blog_do-not-commit.html")
```

## Preface

With EGU 2020 cancelled, this blog post elaborates on the abstract submitted by the development team of the 52°North software project `sos4R`.
See the programme entry for the abstract [EGU2020-19453 in the official programme](https://meetingorganizer.copernicus.org/EGU2020/EGU2020-19453.html).
Please refer to this work by citing the DOI [`https://doi.org/10.5194/egusphere-egu2020-19453`](https://doi.org/10.5194/egusphere-egu2020-19453).

## Introduction

Time series data of in-situ measurements is the key to many environmental studies.
The first challenge in any analysis typically arises when the data needs to be imported into the analysis framework.
Standardisation is one way to lower this burden.
Unfortunate`ly, relevant interoperability standards might be challenging for non-IT experts as long as they are not dealt with behind the scenes of a client application.
One standard to provide access to environmental time series data is the Sensor Observation Service ([SOS](https://en.wikipedia.org/wiki/Sensor_Observation_Service)) specification published by the Open Geospatial Consortium ([OGC](https://www.ogc.org/)). SOS instances are currently used in a broad range of applications such as hydrology, air quality monitoring, and ocean sciences. Data sets provided via an SOS interface can be found around the globe from Europe to New Zealand.
  
The R package [sos4R](https://52north.github.io/sos4R/) (Nüst et al., 2011) is an extension package for the R environment for statistical computing and visualization (https://www.r-project.org/), which has been demonstrated a a powerful tools for conducting and communicating geospatial research (cf. Pebesma et al., 2012).

## Access sensor data with sos4R

sos4R comprises a client that can connect to an SOS server.
The user can use it to query data from SOS instances using simple R function calls.
The following example demonstrates some core operations: connect to an SOS, requesting data availability information and sensor metadata, and retrieving the acdtual data:

```{r classic_sos}
library("sos4R")
fluggs <- SOS(url = "https://fluggs.wupperverband.de/sos2/service", binding = "KVP",
              version = "2.0.0")
```

```{r classic_sos_sensor}
sensor2 <- describeSensor(fluggs, sosProcedures(fluggs)[2], outputFormat = "http://www.opengis.net/sensorML/1.0.1")
sensor2
```

```{r classic_sos_data_availability}
fluggs_data_availability <- getDataAvailability(fluggs)
fluggs_data_availability[1:2]
```


```{r classic_sos_data}
fluggs_offerings <- sosOfferings(fluggs)
fluggs_obs_jan_2019 <- getObservation(fluggs, offering = fluggs_offerings[[1]],
               eventTime = sosCreateTime(fluggs, "2019-01-01::2019-01-31"),
               responseFormat = "http://www.opengis.net/om/2.0")
result <- sosResult(fluggs_obs_jan_2019)
summary(result)
```

However, these functions use terms and operations specific to the OGC services and the SOS standard.
They also require some knowledge about working combinations for requests.
Therefore, a new release of sos4R provides a convenience layer for R users to integrate observation data from data access servers compliant with the SOS standard without any knowledge about the underlying technical standards.
This update also features specific consideration of the [OGC SOS 2.0 Hydrology Profile](TODO) and thereby opens up a new scientific domain.

Especially for non-SOS experts, the wrapper functions remove complexity and technical language specific to OGC specifications with their specific terms, e.g., "FOI", or "procedure".
The wrapper function use more generic terms easily accessible for all users, especially without a strong knowledge of the OGC standards of the Sensor Web Enablement (see ["OGC SWE and SOS"](sos4R-vignette-02-ogc-sos.html) vignette for details).
In general, these functions always return an object of class `data.frame`, even if the result is only a list, in which case the `data.frame` has one column.

The following code chunks demonstrate requesting the same data as above, but using the new wrapper functions.

```{r modern_sos_sites}
sites <- sites(sos = fluggs, includePhenomena = TRUE)

library("kableExtra")
scroll_box(kable(sites), height = "200px")
```

```{r modern_sos_phenomena}
phenomena <- phenomena(sos = fluggs, includeTemporalBBox = TRUE, includeSiteId = TRUE)
watertemp_phenomena <- subset(phenomena, phenomenon == "Wassertemperatur")

observationData <- getData(sos = fluggs,
                           sites = subset(sites, Wassertemperatur == TRUE),
                           phenomena = watertemp_phenomena,
                           begin = as.POSIXct("2019-01-01"),
                           end = parsedate::parse_iso_8601("2019-01-31"))
summary(observationData)
```

The result `data.frame` includes additional metadata.

```{r getData_attributes}
attributes(observationData[[3]])
```

As shown in the code, you can filter the results with objects created with other functions.
The returned data can be limited by thematical, spatial, and temporal filters.
Thematical filtering (phenomena) support the values of the previous functions as inputs.
Spatial filters are either sites, or a bounding box.
Temporal filter is a time period during which observations are made.
Without a temporal extent, the used SOS only returns the last measurement.

Furthermore, you can use the included spatial information about the sites and temporal information about the available phenomena and display sites on a map without the user having to touch coordinates or map projections.
For example, the `sites` object is a `SpatialPointsDataFrame` with proper coordinate reference system (CRS).

```{r modern_sos_bbox}
sites@proj4string
sp::bbox(sites)
```

```{r sites_map, fig.width=6}
suppressPackageStartupMessages(library("mapview"))

mapview(sites)
```

_A more detailed description of the function above can be found in sos4R's documentation TODO LINK TO VIGNETTE HERE._

## Integrated data analysis

The abstraction provided by sos4R makes sensor observation data easily accessible across multiple SOS instances.
It seamlessly integrates distributed observations data, i.e., across organisational boundaries, into transparent and reproducible data analysis workflows.

Plot received data as time series:

```{r getData_timeseries_plot, fig.width=6}
suppressPackageStartupMessages(library("xts"))
tsLaaken <- xts(observationData[observationData$siteID == 'Laaken',3],
                observationData[observationData$siteID == 'Laaken',"timestamp"])
names(tsLaaken) <- "Laaken"
plot(x = tsLaaken, main = "Monthly: Extreme max. Temp. (°C)", yaxis.right = FALSE, legend.loc = 'topleft')
```

TODO: integrate with data from another SOS

## Shiny application

TODO

## References

Nüst D., Stasch C., Pebesma E. (2011) Connecting R to the Sensor Web. In: Geertman S., Reinhardt W., Toppen F. (eds) Advancing Geoinformation Science for a Changing World. Lecture Notes in Geoinformation and Cartography, Springer.

Pebesma, E., Nüst, D., & Bivand, R. (2012). The R software environment in reproducible geoscientific research. Eos, Transactions American Geophysical Union, 93(16), 163–163.
