---
title: "sos4R: List of SOSs"
author:
  name: "Daniel Nüst"
  affiliation: Institute for Geoinformatics, University of Münster, Germany.
  email: daniel.nuest@uni-muenster.de
  url: https://www.nordholmen.net
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{sos4R-vignette-99-services}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A list of SOS instances in production.
A listing here does not imply extensive testing with `sos4R`.
Contributions of demos welcome!

Do you **know about an SOS instance not on this list**?
Open an [issue](https://github.com/52North/sos4R/issues) or a pull request.

```{r setup}
library(sos4R)
```

## NOAA Center for Operational Oceanographic Products and Services (CO-OPS)

```{r coops}
ioosdif <- SOS(url = "http://opendap.co-ops.nos.noaa.gov/ioos-dif-sos/SOS",
               binding = "KVP")
ioosdif
#sosContents(ioosdif)
```

## NOAA National Data Buoy Center

```{r ndbc}
ndbc <- SOS(url = "https://sdf.ndbc.noaa.gov/sos/server.php",
            binding = "KVP")
ndbc
#sosContents(ndbc)
```

## WSV PegelOnline

About: https://www.pegelonline.wsv.de/gast/start

```{r pegelonline}
pegel <- SOS(url = "http://pegelonline.wsv.de/webservices/gis/gdi-sos",
                version = "1.0.0")
pegel
sosContents(pegel)
```

## Axiom Data Science Demo SOS

```{r}
axiom <- SOS(url = "http://dev.demo.i52nsos.axiomdatascience.com/sos", binding = "KVP")
axiom
sosContents(axiom)
```

## Belgian Interregional Environment Agency (IRCEL - CELINE)

```{r irceline}
irceline <- SOS(url = "http://geo.irceline.be/sos/service",
                version = sos200_version)
irceline
#sosContents(irceline)
```

## PANGAEA

About: http://sos.dataportals.pangaea.de/

**Loch Etive, Scotland; Scottish Association for Marine Science**

```{r lochetive}
lochetive <- SOS(url = "http://sos.dataportals.pangaea.de/lochEtive_SOS/sos")
lochetive
sosContents(lochetive)
```

**Koljoefjord observatory, Sweden; University of Gothenburg**

```{r koljoefjord}
koljoefjord <- SOS(url = "http://sos.dataportals.pangaea.de/koljoefjord_SOS/sos")
koljoefjord
sosContents(koljoefjord)
#summary(koljoefjord)
```

## Marthas Vineyard Coastal Observatory SOS

```{r mvco}
MVCO_WHOI <- SOS("http://mvcodata.whoi.edu/cgi-bin/sos/oostethys_sos")
sosContents(MVCO_WHOI)
```

## VITO SOS - Antwerp air quality

```{r idea}
idea <- SOS(url = "http://sensorweb.vito.be:8080/IDEA_52nSOSv3.2/sos")
sosContents(idea)
```

## Wupperverband FLUGSS

About: https://fluggs.wupperverband.de/v2p/web/fluggs

```{r fluggs}
fluggs <- SOS(url = "https://fluggs.wupperverband.de/sos2/service", binding = "KVP",
              version = "2.0.0")
sosContents(fluggs)
```

