---
title: "sos4R: List of SOSs"
author:
  name: "Daniel Nüst"
  affiliation: Institute for Geoinformatics, University of Münster, Germany.
  email: daniel.nuest@uni-muenster.de
  url: https://www.nordholmen.net
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{6. List of SOS instances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A list of SOS instances in production.
A listing here does not imply extensive testing with `sos4R`.
Contributions of demos welcome!

Do you **know about an SOS instance not on this list**?
Open an [issue](https://github.com/52North/sos4R/issues) or a pull request.

```{r setup}
library("sos4R")
```

## NOAA Center for Operational Oceanographic Products and Services (CO-OPS)

```{r coops}
ioosdif <- SOS(url = "http://opendap.co-ops.nos.noaa.gov/ioos-dif-sos/SOS",
               binding = "KVP")
ioosdif
#sosContents(ioosdif)
```

## Integrated Ocean Observing System (IOOS)

More information: https://sdf.ndbc.noaa.gov/sos/

This SOS supports **CSV** as a response format.

```{r ioos}
ioos <- SOS(url = "https://sdf.ndbc.noaa.gov/sos/server.php",
              useDCPs = FALSE,
              timeFormat = "%Y-%m-%dT%H:%M:%SZ")
ioos.off <- sosOfferings(ioos)
obs <- getObservation(sos = ioos,
                      offering = sosName(ioos.off[[100]]),
                      responseFormat = "text/csv",
                      observedProperty = sosObservedProperties(ioos.off[[100]])[1])
obs
```

## NOAA National Data Buoy Center

```{r ndbc}
ndbc <- SOS(url = "https://sdf.ndbc.noaa.gov/sos/server.php",
            binding = "KVP")
ndbc
#sosContents(ndbc)
```

## WSV PegelOnline

About: https://www.pegelonline.wsv.de/gast/start

```{r pegelonline}
pegel <- SOS(url = "http://pegelonline.wsv.de/webservices/gis/gdi-sos",
                version = "1.0.0")
pegel
sosContents(pegel)
```

## Axiom Data Science Demo SOS

```{r}
axiom <- SOS(url = "http://dev.demo.i52nsos.axiomdatascience.com/sos", binding = "KVP")
axiom
sosContents(axiom)
```

## Belgian Interregional Environment Agency (IRCEL - CELINE)

```{r irceline}
irceline <- SOS(url = "http://geo.irceline.be/sos/service",
                version = sos200_version)
irceline
#sosContents(irceline)
```

## PANGAEA

About: http://sos.dataportals.pangaea.de/

**Loch Etive, Scotland; Scottish Association for Marine Science**

```{r lochetive}
lochetive <- SOS(url = "http://sos.dataportals.pangaea.de/lochEtive_SOS/sos")
lochetive
sosContents(lochetive)
```

**Koljoefjord observatory, Sweden; University of Gothenburg**

```{r koljoefjord}
koljoefjord <- SOS(url = "http://sos.dataportals.pangaea.de/koljoefjord_SOS/sos")
koljoefjord
sosContents(koljoefjord)
#summary(koljoefjord)
```

## Marthas Vineyard Coastal Observatory SOS

```{r mvco}
MVCO_WHOI <- SOS("http://mvcodata.whoi.edu/cgi-bin/sos/oostethys_sos")
sosContents(MVCO_WHOI)
```

## Wupperverband FLUGSS

About: https://fluggs.wupperverband.de/v2p/web/fluggs

```{r fluggs}
fluggs <- SOS(url = "https://fluggs.wupperverband.de/sos2/service", binding = "KVP",
              version = "2.0.0")
sosContents(fluggs)
```

## MONALISA

SOS by [MONALISA](http://www.monalisa-project.eu/en/home/Pages/default.aspx) project, operated by [eurac research](http://www.eurac.edu/).

About: http://monalisasos.eurac.edu/sos/

```{r monalisa}
monalisa <- SOS(url = "http://monalisasos.eurac.edu/sos/service", binding = "KVP", version = "2.0.0")
sosContents(monalisa)
```

## NIWA

Two SOS by New Zealands [National Institute of Water and Atmospheric Research](https://www.niwa.co.nz/about/our-company) (NIWA) for hydrometric and climate data, see https://teamwork.niwa.co.nz/display/NEDA/Access+to+Sensor+Data.

```{r niwa_hydro}
niwaHydro <- SOS(url = "https://hydro-sos.niwa.co.nz/",
                  binding = "POX", useDCPs = FALSE,
                  #binding = "KVP",
                  version = "2.0.0")
sosContents(niwaHydro)
```

```{r niwa_hydro_data}
#sosResponseFormats(niwaHydro)
discharge29808 <- getObservation(sos = niwaHydro,
                                 observedProperty = list("Discharge"), # phenomena
                                 featureOfInterest = list("29808"), # sites
                                 #saveOriginal = "../tests/responses/hydro-sos.niwa.co.nz_Discharge_29808.xml",
                                 #responseFormat = "http://www.opengis.net/waterml/2.0",
                                 #inspect = TRUE,
                                 retrieveFOI = FALSE)
sosResult(discharge29808)
```

```{r niwa_climate}
# service offline - HTTP 503
#niwaClimate <- SOS(url = "http://clidb-sos.niwa.co.nz/sos/kvp", binding = "KVP", useDCPs = FALSE, version = "2.0.0")
#sosContents(niwaClimate)
```

## ncSOS

### UCAR unidata

[https://www.unidata.ucar.edu/software/tds/current/reference/Services.html#SOS](https://www.unidata.ucar.edu/software/tds/current/reference/Services.html#SOS)

```{r unidata}
# TODO
```

### Applied Science Associates

[https://github.com/asascience-open/ncSOS/](https://github.com/asascience-open/ncSOS/)

```{r asascience}
# MARACOOS, http://sos.maracoos.org/dev/catalog.html?dataset=wflow700-agg
# http://sos.maracoos.org/dev/sos/wflow700-agg.ncml

# http://sos.maracoos.org/pre/catalog/nerrs_ag/soschwq/catalog.html?dataset=nerrs_ag/soschwq/soschwq.ncml
#http://sos.maracoos.org/pre/sos/nerrs_ag/soschwq/soschwq.ncml

# http://sos.maracoos.org/thredds/catalog/mddnr_ag/Jug_Bay/catalog.html?dataset=mddnr_ag/Jug_Bay/Jug_Bay.ncml
# Capabilities document: http://sos.maracoos.org/stable/sos/mddnr_ag/Jug_Bay/Jug_Bay.ncml
jugbay <- SOS(url = "http://sos.maracoos.org/stable/sos/mddnr_ag/Jug_Bay/Jug_Bay.ncml", binding = "KVP")
jugbay
sosContents(jugbay)

jugbay_observations <- getObservation(jugbay,
                                      offering = sosOfferings(jugbay)[[1]],
                                      procedure = sosProcedures(jugbay)[[1]],
                                      eventTime = sosCreateTime(jugbay, "2020-01-01::2020-01-31"))
str(jugbay_observations)
```
