---
title: "sos4R: New Wrapper Functions for Easier SOS Access"
date: "2019-09-02"
output:
   beamer_presentation:
    includes:
      in_header: "./talk-style.tex"
classoption: "aspectratio=169,8pt"
always_allow_html: yes
---

```{r, include = FALSE}
str(knitr::opts_chunk$get())

knitr::opts_chunk$set(
  comment="",
  collapse = FALSE,
  echo = TRUE,
  cache = TRUE)
```

# Overview

## getting started - SOS4R

R is a statistical tool and programming language tailored for data analysis including spatial data.

It allows to **query data** from standard conform **SOS instances** using simple **R function calls** and does no require any knowledge about the Sensor Web. It is easily extendible for new data models and **opens** the huge amount of analysis and visualization **features of the R environment for the Sensor Web**.
https://52north.github.io/sos4R/

`sos4R` includes a collection of convenience functions which wrap the complex SOS interface with its specific terms (e.g. FOI, procedure).

The wrapper function use more generic terms easily accessible for all users, especially without a strong knowledge of the OGC standards of the Sensor Web Enablement (see ["OGC SWE and SOS"](sos4R-vignette-02-ogc-sos.html) vignette for details).

## SOS4R - project facts

In a [nutshell](https://www.openhub.net/p/sos4R) , sos4R...

- has had 589 commits made by 9 contributors representing 12,680 lines of code
- is mostly written in R with a well-commented source code
- has a well established, mature codebase maintained by a average size development team with increasing Y-O-Y commits
- took an estimated 13 years of effort (COCOMO model) starting with its first commit in May, 2013 ending with its most recent commit about three weeks ago

- CRAN: https://cran.r-project.org/package=sos4R

Sponsors

- NIWA: National Institute of Water and Atmospheric Research, New Zealand
- MuDak-WRM: Multidisciplinary data acquisition as the key for a globally applicable water resource management, funded by BMBF
- TaMIS: Talsperren-Mess-Informations-System, funded by BMBF
- ifgi: student contributions, initial version by Daniel Nüst

<!-- Überblick SOS-Sprech zu Convenience? -->

## sos4R - convenience API

  The key functions of the convenience API are:
  
  - **SOS()** initializes the connection with an SOS endpoint
  - **phenomena()** returns a list of phenomena that the SOS provides. It can also provide details on the recording period and the stations providing this phenomenon
  - **sites()** retrieves details on the feature of interests/measurement stations. This can include their locations, observed phenomena and temporal extent of each time series.
  - **getData()** fetches the desired time series data from the SOS.

The convenience API wraps and combines the default API reusing the available functions in sos4R.

# Phenomena

## Setting up the SOS

```{r setup}
library("sos4R")

niwaHydro <- SOS(url = "https://climate-sos.niwa.co.nz/",
                 binding = "KVP", 
                 useDCPs = FALSE, 
                 version = "2.0.0")
```

## Phenomena{.allowframebreaks}

The function `phenomena(..)` provides information about observed phenomena and time periods of data.

```{r phenomena}
phenomena <- phenomena(sos = niwaHydro)
str(phenomena)
```

```{r phenomena_table}
head(phenomena)
```

\framebreak 

The retrieved data can be extended by time intervals and site identifier for which data is available.

```{r phenomena_include_time}
head(phenomena(sos = niwaHydro, includeTemporalBBox = TRUE))
```

\framebreak 

```{r phenomena_include_siteid}
head(phenomena(sos = niwaHydro, includeSiteId = TRUE))
```

\framebreak 

One can also add both temporal extent and sites.

```{r phenomena_include_all}
head(phenomena(sos = niwaHydro, includeTemporalBBox = TRUE, includeSiteId = TRUE))
```

# Sites

## List sites {.allowframebreaks}

The function `sites(..)` provides information about sites where observations are made, including metadata about the sites (e.g. location).
The returned object is a `SpatialPointsDataFrame`.

```{r sites}
sites <- sites(sos = niwaHydro)
head(sites)
```

<!-- \framebreak -->

<!-- To see all sites, even the ones without any data, use the `empty` parameter. -->

<!-- ```{r sites_empty, eval=TRUE} -->
<!-- sitesEmpty <- sites(sos = niwaHydro, empty = TRUE) -->
<!-- as.data.frame(sitesEmpty)[10:13,] -->
<!-- ``` -->

\framebreak

One can retrieve additional metadata about the phenomena and the time period for which data is available.
Including temporal extent implies inclusion of phenomena.
In the next chunks the object is coerced to a `date.frame` to get a tabular view.

```{r sites_metadata_phen}
sitesPhen <- sites(sos = niwaHydro, includePhenomena = TRUE)
head(colnames(sitesPhen@data))

colnames(sitesPhen@data)[c(1,3,80)] <- c("ID", "FOG", "RAIN")
```

\framebreak

```{r sites_metadata_phen_2}
sitesPhen[,c(1,3,80)]
```

\framebreak

```{r sites_metadata_temporal}
sitesTempBB <- sites(sos = niwaHydro, 
                     includeTemporalBBox = TRUE, includePhenomena = TRUE)
str(sitesTempBB[1:3,c(1,3,80)]@data)
```

## Filter Sites {.allowframebreaks}

One can filter sites using phenomena and temporal extent.

```{r sites_phenomena}
head(sites(sos = niwaHydro, phenomena = phenomena[3,]))
```

\framebreak

```{r sites_time}
head(sites(sos = niwaHydro, 
           begin = as.POSIXct("1904-01-01"), 
           end = as.POSIXct("1905-12-31")))
```

\framebreak

The `SpatialPointsDataFrame` allows access to coordinates with coordinate reference system (CRS).

```{r sites_coords}
library(sp)
coordinates(sites)[1:3,]
```

```{r sites_projection}
sites@proj4string
```

```{r sites_bbox}
bbox(sites)
```

\framebreak

One way of plotting and exploring the sites is:

```{r sites_map, fig.height=4}
library("mapview")
mapview(sites[-16,], legend=FALSE, col.regions="#65c6e4")
```

## Sites as List

The function `siteList(..)` is an analogue, but the result is a list rather than a spatial object.

```{r siteList}
siteList <- siteList(sos = niwaHydro)
str(siteList)
head(siteList)
```

<!-- \framebreak -->

<!-- One can extend the information returned with these parameters: -->

<!-- - `empty` to also show sites without data -->
<!-- - `includePhenomena` to add phenomena to the table (boolean) -->
<!-- - `includeTemporalBBox` to also show the time when data is available (boolean) -->

<!-- ```{r siteList_empty, eval=TRUE} -->
<!-- head(siteList(sos = niwaHydro, empty = TRUE)) -->
<!-- ``` -->

<!-- \framebreak -->

<!-- One can reduce the results with these parameters: -->

<!-- - `phenomena` is a vector of phenomena which must be measured at the sites -->
<!-- - `begin` and `end` define a time interval (date and time class objects) for which some data must be available (sites may have data outside the given interval) -->

<!-- ```{r siteList_phenomena_time} -->
<!-- siteList(sos = niwaHydro, -->
<!--          phenomena = phenomena$phenomenon[80], -->
<!--          begin = as.POSIXct("1950-01-01"),  -->
<!--          end = as.POSIXct("1960-12-31")) -->
<!-- ``` -->

# getData

## Data download {.allowframebreaks}

The function `getData(..)` retrieves data and returns them in a long form 'data.frame' that is ready-to-use for the [`xts`](https://cran.r-project.org/package=xts) or [`spacetime`](https://cran.r-project.org/package=spacetime) package.

The returned data can be limited by thematic, spatial, and temporal filters.
Thematic filtering (phenomena) support the values of the previous functions as inputs.
Spatial filters are either sites, or a bounding box.
Temporal filter is a time period during which observations are made.

Without a temporal extent, the used SOS only returns the last measurement.
\footnotesize
```{r getData_single}
obsData <- getData(sos = niwaHydro,
                   phenomena = phenomena[18,1],
                   sites = siteList[1:2,1]
)
str(obsData,1)
```
\normalsize
\framebreak 

The result `data.frame` includes additional metadata.

```{r getData_attributes}
attributes(obsData[[3]])
```

Request more data with a temporal extent for all sites.

<!-- **Note:** When creating time objects from strings, `as.POSIXct(..)` by default recognises the local time zone and creates a time object in UTC. -->

```{r getData_temporal}
obsData <- getData(sos = niwaHydro,
                   phenomena = phenomena[18,1],
                   sites = siteList[1,1],
                   begin = parsedate::parse_iso_8601("1970-01-01T12:00:00+12:00"),
                   end = parsedate::parse_iso_8601(Sys.time()))
```
\footnotesize
```{r getData_temporal_2}
str(obsData, 2)
```
\normalsize

# Plotting and Analytics

## Time Series Plotting {.allowframebreaks}

Plot received data as time series:

```{r getData_timeseries_plot, fig.width=6, fig.height=3.2}
library(xts)
ts1056 <- xts(obsData[obsData$siteID == '1056',3],
              obsData[obsData$siteID == '1056', "timestamp"])
plot(x = ts1056, 
     main = "Monthly: Extreme max. Temp. (°C)")
```

## Analytics {.allowframebreaks}

Conversion of the time series:

```{r timeseries_decompose, results="hide", warning=FALSE, message=FALSE}
extTs <- merge(ts1056, as.xts(ts(start = c(1981,10), end = c(2018, 8), frequency = 12)))
smplTs <- ts(as.numeric(extTs[,1]), c(1981, 10), frequency = 12)
smplTsFill <- imputeTS::na.interpolation(smplTs)
decTs <- decompose(smplTsFill)
```
\framebreak

```{r timeseries_decompose_2, fig.height=5}
plot(decTs)
lm(y~x, data.frame(y=as.numeric(decTs$trend), x=1:length(decTs$trend)))
```

```{r timeseries_decompose_3, fig.height=5}
plot(decTs$trend,xlim=c(1981, 2020), main="Simple linear trend of monthly maxima")
abline(22.96-(1981+9/12)*0.0022*12, 0.0022*12, col="red")
```

## Multiple Time Series {.allowframebreaks}

You can also retrieve data for phenomena from multiple sites.

```{r getData_multipleB, fig.width=6, fig.height=3.2}
multipleSites <- siteList$siteID[35:38]
obsData <- getData(sos = niwaHydro,
                   phenomena = phenomena[18,1],
                   sites = sites[1:2,]$siteID, # siteList[1:2,1]
                   begin = parsedate::parse_iso_8601(Sys.time()-5.5*365*24*60*60),
                   end = parsedate::parse_iso_8601(Sys.time())
)


ts1056 <- xts(obsData[obsData$siteID == '1056',3],
              obsData[obsData$siteID == '1056',"timestamp"])
names(ts1056) <- "Station__1056"
ts11234 <- xts(obsData[obsData$siteID == '11234',3],
               obsData[obsData$siteID == '11234',"timestamp"])
names(ts11234) <- "Station_11234"
p <- plot(x = na.fill(merge(ts1056, ts11234), list(NA, "extend", NA)), 
     main = "Monthly: Extreme max. Temp. (°C)", 
     ylim=c(14,41))
addLegend("topleft", ncol=2, col = c("black", "red"), lty=1)
plot(p)
```

\framebreak

```{r, fig.height=5}
library(sp)
mapview(sites[1:2,], map.types="OpenTopoMap")
```

